---
output:
  github_document:
    keep_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

# comexstatr

<!-- badges: start -->

[![Lifecycle:
experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN
status](https://www.r-pkg.org/badges/version/comexstatr)](https://CRAN.R-project.org/package=comexstatr)
<!-- badges: end -->

The goal of comexstatr is to make it easy to download, process, and
analyze Brazilian foreign trade statistics, available through the web
app <http://comexstat.mdic.gov.br/>, using the underlying bulk data
<https://www.gov.br/produtividade-e-comercio-exterior/pt-br/assuntos/comercio-exterior/estatisticas/base-de-dados-bruta>.

## Installation

```{r}
##devtools::install_github("leoniedu/comexstatr")
##install.packages("tictoc")
```

###If you have problems installing arrow, see:

https://arrow.apache.org/docs/r/articles/install.html#:~:text=Step%201%20-%20Using%20a%20computer%20with%20internet,created%20my_arrow_pkg.tar.gz%20to%20the%20computer%20without%20internet%20access


## Examples

``` {r}
library(comexstatr)
library(dplyr)
library(tictoc)

##downloading
tic()
comexstat_download(
  ##force_download = TRUE
  )
## might need something like this if you get ssl errors. 
##comexstat_download(method="wget", extra="--no-check-certificate")
toc()


```

### Main trade partners, treating countries in Mercosul and European Union as blocks.

```{r}

msul <- pais_bloco()%>%
  filter(co_bloco==111)%>%
  pull(co_pais)
eu <- pais_bloco()%>%
  filter(co_bloco==22)%>%
  pull(co_pais)

pb <- pais()%>%
  transmute(co_pais, 
            partner=
              case_when(co_pais%in%msul ~ "Mercosul",
                        co_pais%in%eu ~ "European Union",
                        TRUE ~ no_pais_ing)
              )

cstat_top_0 <- comexstat()|>
  left_join(pb) |> 
  #filter(co_ano>=2017)|>
  group_by(partner)|>
  summarise(vl_fob=sum(vl_fob))|>
  ungroup() |> 
  arrange(desc(vl_fob))|>
  collect()|>
  slice(1:5)

cstat_top <- comexstat() |>
  left_join(pb) |> 
  #filter(co_ano>=2017)%>%
  semi_join(cstat_top_0, by=c("partner"))%>%
  group_by(co_ano, partner, fluxo)|>
  summarise(vl_fob=sum(vl_fob))%>%
  collect

library(ggplot2)
qplot(co_ano, vl_fob_bi, data=cstat_top%>%filter(co_ano<2022)%>%mutate(vl_fob_bi=vl_fob/1e9), color=partner, geom="line") +
  facet_wrap(~fluxo) +
  labs(color="", x="", y="US$ Bi (FOB)") +
  theme_linedraw() + theme(legend.position="bottom")

```



## By ISIC - International Standard Industrial Classification of All Economic Activities


```{r}

isic <- comexstat() |> 
  left_join(ncms()%>%select(co_ncm, isic=no_isic_secao_ing))%>%
  group_by(isic, co_ano, fluxo)%>%
  summarise(vl_fob=sum(vl_fob))%>%
  collect



qplot(co_ano, vl_fob_bi, 
      data=isic%>%mutate(vl_fob_bi=vl_fob/1e9), color=isic, geom="line") +
  facet_wrap(~fluxo) +
  labs(color="", x="", y="US$ Bi (FOB)") +
  theme_linedraw() + theme(legend.position="bottom")

```




### By state

```{r}

bystate <- comexstat() |> 
  group_by(state=sg_uf_ncm, co_ano, fluxo)%>%
  summarise(vl_fob=sum(vl_fob))%>%
  collect

topstate <- bystate%>%group_by(state)%>%summarise(vl_fob=sum(vl_fob))%>%arrange(-vl_fob)%>%head(3)



qplot(co_ano, vl_fob_bi, 
      data=bystate%>%
        semi_join(topstate, by="state")%>%
        mutate(vl_fob_bi=vl_fob/1e9), color=state, geom="line") +
  facet_wrap(~fluxo) +
  labs(color="", x="", y="US$ Bi (FOB)") +
  theme_linedraw() + theme(legend.position="bottom")
```




